Finally Tagless, Partially Evaluated

### 2007-07-27 tagless/tagless-final.tex
Extend related work

Our initial motivation came from several papers that justify advanced
type systems, in particular GADTs, by embedded interpreters
\citep{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple} and
CPS transformations \citep{Guillemette-Monier-PLPV,chen-typeful}.
We admire all this technical machinery, but
these motivating examples do not need it.
Although GADTs may indeed be more flexible and easier to use, they are
unavailable in mainstream ML and implemented problematically
in GHC currently. We also wanted to find the minimal
set of widespread language features needed for tagless
type-preserving interpretation.

@@

Our partial evaluator establishes a bijection |repr_pe| between static
and dynamic types (the valid values of |'sv| and |'dv|), and between
static and dynamic terms.  It is customary to implement such a bijection
using an injection\hyp projection pair, as done for interpreters
\ifshort \cite{Ramsey-ML-module-mania,Benton-embedded-interpreters}\else
by \citet{Ramsey-ML-module-mania} and \citet{Benton-embedded-interpreters}\fi,
partial evaluation \ifshort \cite{Danvy-TDPE}\else by \citet{Danvy-TDPE}\fi,
and type-level functions \ifshort \cite{oliveira-typecase}\else by
\citet{oliveira-typecase}\fi.  As explained in~\S\ref{S:PE-solution}, we
avoid injection and projection at the type level by adding an argument
to |repr|.
\begin{comment}
Our solution could have been even more straightforward if we had
total type-level functions at our disposal (|repr_pe| in
\S\ref{S:PE-solution}) in MetaOCaml.  Simple type-level computations 
certainly ought to become mainstream.
\end{comment}

@@

>>>>

Our initial motivation came from several papers that justify advanced
type systems, in particular GADTs, by embedded interpreters
\citep{WalidICFP02,taha-tag,xi-guarded,peyton-jones-simple} and
CPS transformations \citep{Guillemette-Monier-PLPV,shao-type-toplas,chen-typeful}.
We admire all this technical machinery, but
these motivating examples do not need it.
Although GADTs may indeed be more flexible and easier to use, they are
unavailable in mainstream ML and implemented problematically
in GHC currently. We also wanted to find the minimal
set of widespread language features needed for tagless
type-preserving interpretation.

@@

Our partial evaluator establishes a bijection |repr_pe| between static
and dynamic types (the valid values of |'sv| and |'dv|), and between
static and dynamic terms.  It is customary to implement such a bijection
using an injection\hyp projection pair, as done for interpreters
\ifshort \cite{Ramsey-ML-module-mania,Benton-embedded-interpreters}\else
by \citet{Ramsey-ML-module-mania} and \citet{Benton-embedded-interpreters}\fi,
partial evaluation \ifshort \cite{Danvy-TDPE}\else by \citet{Danvy-TDPE}\fi,
and type-level functions \ifshort \cite{oliveira-typecase}\else by
\citet{oliveira-typecase}\fi.  As explained in~\S\ref{S:PE-solution}, we
avoid injection and projection at the type level by adding an argument
to |repr|.
Our solution could have been even more straightforward if MetaOCaml
provided total type-level functions such as |repr_pe| in
\S\ref{S:PE-solution}---simple type-level computations 
ought to become mainstream.

@@

Using Haskell, \citet{Guillemette-Monier-PLPV} implement a CPS transformation
for HOAS terms and statically assure that it preserves object types.
They represent proofs of type preservation as terms of a GADT, which is not
sound (as they admit in \S4.2) without a separate totality check because
any type is trivially inhabited by a nonterminating term in Haskell.
In contrast, our CPS transformations use simpler types than GADTs and
assure type preservation at the (terminating) type level rather than
the term level of the metalanguage.
\Citeauthor{Guillemette-Monier-PLPV} review other
type\hyp preserving CPS transformations (mainly in the context of typed intermediate
languages), in particular
\citets{shao-type-toplas} and \citets{chen-typeful}.
These approaches use de Bruijn indices and fancier
type systems with type-level functions, GADTs, or type\hyp equality
proofs.
